---
title: "Assignment Week 7"
format: html
editor: visual
---

## Group 04:

-   Edir Sebastian Vidal Castro (s243564),
-   Morten Orebo Holmström (s242223),
-   Nithiyashri Jayashankar (s244356),
-   Tilman Johannes Nelissen (s233043),
-   Thea Rehm Rosholm (s225019)

## Libraries used

```{r}
library(tidyverse)
library(broom)  # devtools::install_github("tidymodels/broom")
library(cowplot)
```

## Initializing data

We’ll be analyzing the `biopsy` dataset, which comes originally from the MASS package. It is a breast cancer dataset from the University of Wisconsin Hospitals, Madison from Dr. William H. Wolberg. He assessed biopsies of breast tumors for 699 patients; each of nine attributes was scored on a scale of 1 to 10. The true outcome (benign/malignant) is also known.

```{r}
biopsy <- read_csv("https://wilkelab.org/classes/SDS348/data_sets/biopsy.csv")
```

## Look at the data in PC coordinates

```{r}
pca_fit <- biopsy |>
  select(where(is.numeric)) |> # retain only numeric columns
  scale() |>
  prcomp(scale = TRUE) # do PCA on scaled data
```

```{r}
pca_fit |>
  augment(biopsy) |> # add original dataset back in
  ggplot(aes(.fittedPC1, .fittedPC2, color = outcome)) + 
  geom_point(size = 1.5) +
  scale_color_manual(
    values = c(malignant = "#D55E00", benign = "#0072B2")
  ) +
  theme_half_open(12) + background_grid()
```

## Rotation matrix

```{r}
# extract rotation matrix
pca_fit |>
  tidy(matrix = "rotation")
```

```{r}
# define arrow style for plotting
arrow_style <- arrow(
  angle = 20, ends = "first", type = "closed", length = grid::unit(8, "pt")
)

# plot rotation matrix
pca_fit |>
  tidy(matrix = "rotation") |>
  pivot_wider(names_from = "PC", names_prefix = "PC", values_from = "value") |>
  ggplot(aes(PC1, PC2)) +
  geom_segment(xend = 0, yend = 0, arrow = arrow_style) +
  geom_text(
    aes(label = column),
    hjust = 1, nudge_x = -0.02, 
    color = "#904C2F"
  ) +
  xlim(-1.25, .5) + ylim(-.5, 1) +
  coord_fixed() + # fix aspect ratio to 1:1
  theme_minimal_grid(12)
```

## Look at the variance explained by each PC

```{r}
pca_fit |>
  tidy(matrix = "eigenvalues")
```

```{r}
pca_fit |>
  tidy(matrix = "eigenvalues") |>
  ggplot(aes(PC, percent)) +
  geom_col(fill = "#56B4E9", alpha = 0.8) +
  scale_x_continuous(breaks = 1:9) +
  scale_y_continuous(
    labels = scales::percent_format(),
    expand = expansion(mult = c(0, 0.01))
  ) +
  theme_minimal_hgrid(12)
```
